Scoring Patients

### 2017-02-24 Patient Report.Rmd
Updating reports to use with new data

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "15 November 2016"
output: word_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)

library(tibble)

@@

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient', sep='/')

@@

sum.patient.data = ddply(patient.info, .(Patient, Status), summarise, years=paste(range(Endoscopy.Year), collapse='-'), 
      n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T))

@@

## Gamma = 250
```{r cors, echo=F, message=T, warning=F, fig.height=6, fig.width=6, fig.align='left'}
## Cluster the samples by coverage then pull out regions

@@

@@

  # ORDER segvals by endoscopy year etc
  cur.pt = subset(patient.info, Patient == patient.name)
  segvals[, c(colnames(segvals[1:5]), cur.pt$Samplename)]

@@

      variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
      variable.regions = segvals[variable.region.indices, patient.data[[patient.name]]$info$Samplename] 
      patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )

@@

      gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')
      print(gg3)
      #pt.plots[[patient.name]][['norm.hist']] = gg

@@

```{r scores, echo=F, message=T, warning=F, fig.height=4,fig.width=4,fig.align='left'} 
#pander(cor.scores[order(cor.scores$Status, cor.scores$similar),], justify='left')

#ggplot(cor.scores, aes(x=similar,y=different, color=Status)) + geom_point()

#ggplot(cor.scores, aes(y=abs(similar+different), x=c(1:nrow(cor.scores)), color=Status)) + geom_point()
```

## Variable Regions
```{r var_reg, echo=F, message=F, warning=F}
rm(prog.regions, np.regions)
for (patient.name in sum.patient.data$Patient) {
  patient = patient.data[[patient.name]]

@@

t.test(subset(np.regions, chrom %in% c(1:12))$chr.ratio, subset(np.regions, chrom %in% c(13:22))$chr.ratio)

t.test(subset(prog.regions, chrom %in% c(1:12))$chr.ratio, subset(prog.regions, chrom %in% c(13:22))$chr.ratio)

@@

There appears to be no enrichment based on the longer vs shorter chromosomes either (fisher `r ftL$p.value`)

```{r cor, echo=F, message=T, warning=F, fig.height=4,fig.width=4, fig.align='left'}

@@

  sample.normalised.segvals = patient.data[[name]]$norm.seg.vals # prog
  sample.normalised.segvals =  sample.normalised.segvals[, patient.data[[name]]$info$Samplename ]

  colnames(sample.normalised.segvals) =  paste(patient.data[[name]]$info$Endoscopy.Year,  c(1:nrow(patient.data[[name]]$info)) , sep='.')

@@

## Look at Cancer Consensus Genes in the variable regions

```{r, hmm, echo=F, message=F, warning=F}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)

@@

m = melt(gene.p[1:50,c(1:4)], id.vars=c('Gene.Symbols', 'CN'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
m$CN = factor(m$CN)

p = ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')
p + geom_point(data=m[m$CN == 'gain' & m$value > 0,], aes(label='gain',y=Gene.Symbols, x=variable), size=2, shape=24, fill='red', color='red') +
  geom_point(data=m[m$CN == 'loss' & m$value > 0,], aes(label='loss',y=Gene.Symbols, x=variable), color='green4', size=2, shape=25, fill='green4')

gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
m = melt(gc[1:50,], id.vars=c('Gene.Symbols'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')

@@

## Is HGD over represented for variable regions?
```{r, hgd, echo=F, message=F, warning=F}

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  print(name)
  sn = grep('HGD', patient.data[[name]]$info$Pathology)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next
  samplenames = patient.data[[name]]$info$Samplename

@@

pander(pvnp, justify='left', 
       caption=paste("Variable regions found in HGD vs non-HGD samples of progessor patients.  There's no significant difference between them (t-test p=",signif(tt$p.value,2),').  Not all patients have HGD samples (NAs).',sep=''))

>>>>

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "15 November 2016"
output:
  html_document: 
    toc: yes
  word_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(tibble)

@@

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient-OLD', sep='/')

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = subset(sum.patient.data, Initial.Analysis)

patient.info = subset(patient.info, Patient %in% sum.patient.data$Patient & Batch.Name %in% levels(patient.info$Batch.Name)[1:5])

@@

```{r cors, echo=F, message=T, warning=F, fig.height=6, fig.width=6, fig.align='left', include=F}
## Cluster the samples by coverage then pull out regions

@@

  if(!file.exists(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""))) next

@@

  # ORDER segvals by endoscopy year etc
  cur.pt = subset(patient.info, Patient == patient.name)
  segvals = segvals[, c(colnames(segvals[1:5]), intersect(colnames(segvals), cur.pt$Samplename))]

@@

      variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
      variable.regions = segvals[variable.region.indices, intersect(patient.data[[patient.name]]$info$Samplename, colnames(segvals))] 
      patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )

@@

      gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')
     # print(gg3)
      #pt.plots[[patient.name]][['norm.hist']] = gg

@@

# Data

This data was generated in the following pipeline:

* Alignment using QDNAseq
* Filter excluded regions
* Joint segmentation per patient across all patient-derived-samples using PCF  (copynumber::multipcf method)

The result looks like is two data matrices:

1. Segmented values 
```{r}
head(patient.data[['AD0258']]$seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[['AD0258']]$seg.vals[,-(1:5)]), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.03) + labs(title='AD0258 segmented value distribution', x='', y='')
```

2. Row (e.g. per-region across timepoints) normalized values from the matrix above
```{r}
head(patient.data[['AD0258']]$norm.seg.vals)
```

```{r echo=F, warning=F, message=F, fig.height=10, fig.width=10}
ggplot( melt(patient.data[['AD0258']]$norm.seg.vals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + labs(title='AD0258 normalized segmented value distribution', x='', y='')

## Per-sample normalization
#nsv = patient.data[['AD0258']]$norm.seg.vals
#nsv[] = lapply(patient.data[['AD0258']]$seg.vals[,-(1:5)], function(x) (x-mean(x))/sd(x))

#ggplot(melt(nsv), aes(x=value))  + 
#        facet_wrap(~variable,scales = "free_x") + 
#        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + labs(title='AD0258 normalized segmented value distribution', x='', y='')

Patients do not share the same chromosomal regions, though they can be overlapped, or the same timepoints. So all analyses have been conducted on a per-patient basis.

```{r general, warning=F, echo=F, message=F}
gen = do.call(rbind, lapply(patient.data, function(x) nrow(x[['seg.vals']])))
colnames(gen) = c('Number.Regions')

sum.patient.data = subset(sum.patient.data, Patient %in% names(patient.data))

sum.patient.data = merge(sum.patient.data, gen, by.x='Patient', by.y='row.names')
sum.patient.data = arrange(sum.patient.data, Status, total.samples)
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))

pander(sum.patient.data[,c('Patient','Status','years','total.samples','highest.path','Number.Regions')], justify='left')
```

# Segmented Value Distributions
```{r echo=F, message=F, warning=F, fig.height=15, fig.width=10}

lapply(patient.data, function(pt) {
  name = pt$info$Patient[1]
  status = pt$info$Status[1]

  g1 = ggplot( melt(pt[['seg.vals']][,-(1:5)]), aes(x=value) ) + 
      facet_wrap(~variable,scales = "free_x") + 
      geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.03) + 
      labs(title='segmented value distribution', x='', y='')

  g2 = ggplot( melt(pt[['norm.seg.vals']]), aes(x=value) ) + 
      facet_wrap(~variable,scales = "free_x") + 
      geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + 
      labs(title='normalized segmented value distribution', x='', y='')

  do.call('grid.arrange', list(g1,g2, top=paste(name, status, sep=" - ")))
})

# Variable Regions

These are regions that are called as "variant" based on cutoffs derived from a single sample that had high-depth (50x) WGS done and was then downsampled. These cutoffs are:

* Standard deviation of the segmented value matrix >= 0.08
* The number of probes used >= 67, covers 1Mb 
* Gains are called at >= 1.1
* Losses are called at <= 0.9

```{r var_reg, echo=F, message=F, warning=F}
rm(prog.regions, np.regions)
for (patient.name in sum.patient.data$Patient) {
  patient = patient.data[[patient.name]]

@@

#t.test(subset(np.regions, chrom %in% c(1:12))$chr.ratio, subset(np.regions, chrom %in% c(13:22))$chr.ratio)

#t.test(subset(prog.regions, chrom %in% c(1:12))$chr.ratio, subset(prog.regions, chrom %in% c(13:22))$chr.ratio)

@@

There appears to be no enrichment based on the length of chromosomes either (fisher `r ftL$p.value`)

```{r cor, echo=F, message=F, warning=F, fig.height=10,fig.width=10, fig.align='left'}

@@

  sample.normalised.segvals = patient.data[[name]]$norm.seg.vals # prog
  sample.normalised.segvals =  sample.normalised.segvals[, intersect(patient.data[[name]]$info$Samplename, colnames(sample.normalised.segvals)) ]

  info = subset(patient.data[[name]]$info, Samplename %in% colnames(sample.normalised.segvals))

  colnames(sample.normalised.segvals) =  paste(info$Endoscopy.Year,  c(1:nrow(info)) , sep='.')

@@

## How do Cancer Gene Consensus genes map to the variable regions?

```{r, hmm, echo=F, message=F, warning=F, fig.height=40, fig.width=10}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)

@@

m = melt(gene.p[,c(1:4)], id.vars=c('Gene.Symbols', 'CN'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
m$CN = factor(m$CN)

p = ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')
p + geom_point(data=m[m$CN == 'gain' & m$value > 0,], aes(y=Gene.Symbols, x=variable), size=2, shape=24, fill='red', color='red') +
  geom_point(data=m[m$CN == 'loss' & m$value > 0,], aes(y=Gene.Symbols, x=variable), color='green4', size=2, shape=25, fill='green4')

gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
m = melt(gc, id.vars=c('Gene.Symbols'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')

@@

# Is HGD over represented for variable regions?

It's possible that the variable regions we find are entirely within the HGD/IMC samples.  So are they overrepresented?

```{r, hgd, echo=F, message=F, warning=F}

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  print(name)
  sn = grep('HGD|IMC', patient.data[[name]]$info$Pathology)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next
  samplenames = patient.data[[name]]$info$Samplename

@@

ct = cor.test(subset(sum.patient.data, Status == 'P')$total.samples, pvnp$hgd.ratio)

pander(pvnp, justify='left', 
       caption=paste("Variable regions found in HGD vs non-HGD samples of progessor patients.  There's no significant difference between them (t-test p=",signif(tt$p.value,2),').  Not all patients have HGD samples (NAs). There is also no significant correlation between the number/ratio of variable regions found in HGD samples vs non (Pearsons p=',signif(ct$p.value, 2),')',  sep=''))
